Stability concerns in random matrices

### 2018-03-28 random_matrix_stability.Rmd
Fix bullet

- I started analysing random communities of simulated species using matrices and a mathematical trick similar to that used in @Patel2018 in attempt to see if variation in generation time among species will stabilise or destabilise large communities.
- I initially found variation in generation time to be highly stabilising, and, in particular, as species richness increased, the overwhelming number of stable random communties were stable *due to variation in generation time* (and would otherwise be unstable). Results are summarised in [this figure](#old_fig_1) and [this table](#old_tab_1).
- I then read the very recent paper by @Dougoud2018, who showed that the classic stability criteria used by @May1972 and @Allesina2012 (among others) have been flawed. Some "stable" communities in these studies are stable (i.e., return to equilibrium values if perturbed) at *negative* species densities. @Dougoud2018 demonstrate how to check whether or not random communities are "feasible", meaning that they are stable and have species densities that are all positive (and hence, sensible ecological communities).
- I applied the criteria of @Dougoud2018 to my random communities under the same parameter values that they and others have used; I also looked at communities characterised only by competitive, mutualistic, and predator-prey interactions.
- Strangely, variation in generation time has the exact *opposite* effect in random, competitive, and predator-prey communities; variation in generation time therefore *increases* stability but *decreases* feasibility (i.e., variation in generation time shifts communities toward being "stable" with negative species densities and away from being "stable" with positive species densities). The overall effect is therefore to make real ecological communities less stable.
- I am still working out what this means conceptually by multiplying each row of the matrix by a single number, the shape of the matrix is somehow being changed to decrease the eigenvalues.
- The exception to this is mutualistic communities -- variation in generation time had zero effect on mutualist community stability.
- **I am tempted to broaden these results to make a more general point about complex dynamic systems**: "Variation in process time increases stability but decreases feasibility in complex systems", with a focus on biological systems. There might be some types of systems where negative values are actually meaningful (e.g., phenotypes in evolving systems, gene regulatory networks? Suggestions very welcome), just not in ecological ones where species densities obviously cannot be negative.

@@

We can delve a bit deeper into this mess by looking at the distribution of eigenvalues in sample communities that are unstable, stable, destabilised, or stabilised. The plot below shows these examples for communities where species number $S = 20$. This gives the same type of plots as shown in @Allesina2012. Black points show eigenvalues before variation in generation time is added, and red points show eigenvalues after variation in generation time is added. The shift is not typically drastic, but sometimes causes the sign of the real components (x-axis) to flip, thereby potentially affecting stability. Immediately below shows plots for a random community; where not points are plot in any of the below, no examples exist.

>>>>

- I started analysing random communities of simulated species using matrices and a mathematical trick similar to that used in @Patel2018 in attempt to see if variation in generation time among species will stabilise or destabilise large communities.
- I initially found variation in generation time to be highly stabilising, and, in particular, as species richness increased, the overwhelming number of stable random communties were stable *due to variation in generation time* (and would otherwise be unstable). Results are summarised in [this figure](#old_fig_1) and [this table](#old_tab_1).
- I then read the very recent paper by @Dougoud2018, who showed that the classic stability criteria used by @May1972 and @Allesina2012 (among others) have been flawed. Some "stable" communities in these studies are stable (i.e., return to equilibrium values if perturbed) at *negative* species densities. @Dougoud2018 demonstrate how to check whether or not random communities are "feasible", meaning that they are stable and have species densities that are all positive (and hence, sensible ecological communities).
- I applied the criteria of @Dougoud2018 to my random communities under the same parameter values that they and others have used; I also looked at communities characterised only by competitive, mutualistic, and predator-prey interactions.
- Strangely, variation in generation time has the exact *opposite* effect in random, competitive, and predator-prey communities; variation in generation time therefore *increases* stability but *decreases* feasibility (i.e., variation in generation time shifts communities toward being "stable" with negative species densities and away from being "stable" with positive species densities). The overall effect is therefore to make real ecological communities less stable.
- I am still working out what this means conceptually (trying to do this by plotting the eigenvalues and comparing different random systems).
- The exception to all of this is mutualistic communities -- variation in generation time had zero effect on mutualist community stability.
- **I am tempted to broaden these results to make a more general point about complex dynamic systems**: "Variation in process time increases stability but decreases feasibility in complex systems", with a focus on biological systems. There might be some types of systems where negative values are actually meaningful (e.g., phenotypes in evolving systems, gene regulatory networks? Suggestions very welcome), just not in ecological ones where species densities obviously cannot be negative.

@@

We can delve a bit deeper into this mess by looking at the distribution of eigenvalues in sample communities that are unstable, stable, destabilised, or stabilised. The plot below shows these examples for communities where species number $S = 20$. This gives the same type of plots as shown in @Allesina2012. Black points show eigenvalues before variation in generation time is added, and red points show eigenvalues after variation in generation time is added. The shift is not typically drastic, but sometimes causes the sign of the real components (x-axis) to flip, thereby potentially affecting stability. Immediately below shows plots for a random community; where no points are plotted in any of the figures below, no examples exist.
